An Efficient Minibatch Acceptance Test for Metropolis-Hastings

### 2016-10-02 nips_2016.tex
Slight updates/clarifications to Appendix A. Still not quite done ...

\subsection{A Worst-Case Gaussian Example}\label{ssec:gaussian_example}
Consider a Gaussian model where $x_1,\ldots,x_N \simiid \mathcal{N}(\theta,1)$
with a known variance $\sigma^2=1$, true mean $\theta=0.5$, and a uniform prior
on $\theta$. The log likelihood ratio is
\begin{equation}
\Lambda^*(\theta,\theta') =  \frac{N}{b}\sum_{i=1}^b \log\frac{p(x_i|\theta')}{p(x_i|\theta)}=
  N(\theta'-\theta)\left(\frac{1}{b}\sum_{i=1}^b x_i-\theta-\frac{\theta'-\theta}{2}\right)
\end{equation}
which is normally distributed over selection of the Normal samples $x_i$.  Since
the $x_i$ have unit variance, their mean has variance $1/b$, and the variance of
$\Lambda^*(\theta,\theta')$ is $\sigma^2(\Lambda^*) = (\theta'-\theta)^2N^2/b$.
In order to pass a hypothesis test that $\Lambda > \psi$, there needs to be a
large enough gap (several $\sigma(\Lambda^*)$) between
$\Lambda^*(\theta,\theta')$ and $\psi(u,\theta,\theta')$.

@@

Similar results can be shown for other distributions and proposals by
identifying regions in product space $(\theta,\theta'-\theta,u)$ such that the
hypothesis test needs to separate nearly-equal values.  It follows that the
accelerated M-H tests in~\cite{cutting_mh_2014,icml2014c1_bardenet14} require at
least a constant fraction $\geq c$ in the amount of data consumed per test
compared to full-dataset tests, so their speed-up is at most $1/c$.

@@

    \item Added more detail for the $\Lambda^*(\theta,\theta') \not\approx
    \psi(u,\theta,\theta')$ part.
\end{enumerate}
}

@@

These events together guarantee that $\Lambda^*(\theta,\theta')$ is negative,
which means that we can find a $u \in [0,1]$ so that $\psi(u,\theta,\theta') =
\log u < 0$ equals $E[\Lambda^*(\theta,\theta')]$.  Specifically, choose $u_0$
to satisfy $\log u_0 = E[\Lambda^*(\theta,\theta')]$.  Using $E[x_i] = 0.5$, we
see that
\begin{align}
    \log u_0 &= N(\theta'-\theta)\frac{1}{b} \cdot E\left[\sum_{i=1}^b x_i-\theta-\frac{\theta'-\theta}{2}\right] \\
    &= -N(\theta'-\theta)\left(\theta-0.5+\frac{\theta'-\theta}{2}\right).
\end{align}
Next, consider the minibatch acceptance test $\Lambda^*(\theta,\theta') \not\approx
\psi(u,\theta,\theta')$, where $\not\approx$ informally means ``significantly
different from'' under the distribution over samples of $x_i$. This turns out to
be
\begin{align}
\Lambda^*(\theta,\theta') \hspace{0.1in} \not\approx \hspace{0.1in}
\psi(u_0,\theta,\theta') &\iff N(\theta'-\theta) \cdot \frac{1}{b}\sum_{i=1}^b x_i-\theta-\frac{\theta'-\theta}{2} \hspace{0.1in} \not\approx \hspace{0.1in} \log u_0 \\
&\iff \frac{1}{b}\sum_{i=1}^b x_i-\left(\theta+\frac{\theta'-\theta}{2} + \frac{\log u_0}{N(\theta'-\theta)}\right) \hspace{0.1in} \not\approx \hspace{0.1in} 0 \\
&\iff \frac{1}{b}\sum_{i=1}^b x_i-0.5 \hspace{0.1in} \not\approx \hspace{0.1in} 0.
\end{align}
Since the $x_i$ have mean 0.5, the resulting test with our chosen $u_0$ will
never correctly succeed and must use all $N$ data points.  Furthermore, if we
sample values of $u$ near enough to $u_0$, the terms in parenthesis will not be
sufficiently different from 0.5 to allow the test to succeed.

{\color{blue} Daniel: TODO I have not yet updated the rest of this section.}

The choices above for $\theta$ and $\theta'$ guarantee that
\begin{equation}
    \log u_0 \in -[0.5,1][0.75,1.5] = -[0.375,1.5]
\end{equation}
and consider the range of $u$ values
\begin{equation}
    \log u \in \log u_0 + [-0.5,0.375]
\end{equation}
the size of the range in $u$ is at least $\exp([-2,-1.125]) = [0.13534,0.32465]$ and occurs
with probability at least $p_2=0.18932$. With $u$ in this range, we rewrite the test as:
\begin{equation}
    \frac{1}{b}\sum_{i=1}^b x_i-0.5 \hspace{0.1in} \not\approx \hspace{0.1in} \frac{\log u/u_0}{N(\theta'-\theta)}
\end{equation}
so that the LHS has expected value zero.  Given our choice of intervals for the
variables, the RHS is in the range $1/\sqrt{N}[-1,0.75]$.  The standard
deviation of the LHS given the interval constraints is at least $0.5/\sqrt{b}$.
And so the gap between LHS and RHS is at most $2\sqrt{b/N}$ standard deviations.

The samples $\theta$, $(\theta'-\theta)$ and $u$ are drawn independently and so
the probability of the conjunction of these events is $c = p_0 p_1 p_2 =
0.0085$.

>>>>

\subsection{A Worst-Case Gaussian Example}\label{ssec:gaussian_example}
Consider a Gaussian model where $x_1,\ldots,x_N \simiid \mathcal{N}(\theta,1)$
with a known variance $\sigma^2=1$, true mean $\theta=0.5$, and a uniform prior
on $\theta$. The log likelihood ratio is
\begin{equation}\label{eq:lemma_ll_ratio}
\Lambda^*(\theta,\theta') =  \frac{N}{b}\sum_{i=1}^b \log\frac{p(x_i|\theta')}{p(x_i|\theta)}=
  N(\theta'-\theta)\left(\frac{1}{b}\sum_{i=1}^b x_i-\theta-\frac{\theta'-\theta}{2}\right)
\end{equation}
which is normally distributed over selection of the Normal samples $x_i$.  Since
the $x_i$ have unit variance, their mean has variance $1/b$, and the variance of
$\Lambda^*(\theta,\theta')$ is $\sigma^2(\Lambda^*) = (\theta'-\theta)^2N^2/b$.
In order to pass a hypothesis test that $\Lambda > \psi$, there needs to be a
large enough gap (several $\sigma(\Lambda^*)$) between
$\Lambda^*(\theta,\theta')$ and $\psi(u,\theta,\theta')$.

@@

Similar results can be shown for other distributions and proposals by
identifying regions in product space $(\theta,\theta',u)$ such that the
hypothesis test needs to separate nearly-equal values.  It follows that the
accelerated M-H tests in~\cite{cutting_mh_2014,icml2014c1_bardenet14} require at
least a constant fraction $\geq c$ in the amount of data consumed per test
compared to full-dataset tests, so their speed-up is at most $1/c$.

@@

    \item Added more detail for the $\Lambda^*(\theta,\theta') \not\approx
    \psi(u,\theta,\theta')$ part.

    \item (TODO later) After Equation~\ref{eq:accept_test_zero}, we say that
    ``the resulting test with our chosen $u_0$ will never correctly succeed
    [...]'', but I think we should say it will not succeed with high
    probability, because we could select a minibatch of samples such that the
    mean \emph{actually is} different from 0.5 (it depends on the distribution
    of the variable that corresponds to $Y = \frac{1}{b}\sum_{i=1}^b x_i$).

    \item (TODO later) Unfortunately, I am unable to understand why ``the size
    of the range in $u$ is at least $\exp([-2,-1.125])$'', unless the real range
    is $\exp([-2,0])$ (as I think it should be) and you were simply cropping off
    the extra range.

    \item (TODO later) I do not understand why after
    Equation~\ref{eq:accept_test_rewritten}, we say that the standard deviation
    of the LHS ``given the interval constraints is at least $0.5/\sqrt{b}$.'' We
    already know that the $X_i$ are iid $\mathcal{N}(\theta = 0.5, 1)$ variables
    with a fixed $\theta = 0.5$ value, so we already know the standard deviation
    is supposed to be $\sqrt{({\rm Var}(x_1) + \cdots + {\rm Var}(x_b))/b^2} =
    1/\sqrt{b}$.

    \item At the end, it said the joint distribution of
    $(\theta,\theta'-\theta,u)$, but I changed it to $(\theta,\theta',u)$, based
    on your change to the text of Lemma~\ref{lem:worst_case}. But here's a
    question: how is the sample of $\theta'$ (or even $\theta'-\theta$, as it
    was written earlier) independent of $\theta$?
\end{enumerate}
}

@@

These events together guarantee that $\Lambda^*(\theta,\theta')$ is negative,
intuitively because the next choice of $\theta'$ must be ``further away'' from
0.5 than $\theta$.  This fact means that we can find a $u \in (0,1)$ so that
$\psi(u,\theta,\theta') = \log u < 0$ equals $E[\Lambda^*(\theta,\theta')]$.
Specifically, choose $u_0$ to satisfy $\log u_0 = E[\Lambda^*(\theta,\theta')]$.
Using $E[x_i] = 0.5$ and Equation~\ref{eq:lemma_ll_ratio}, we see that
\begin{align}
    \log u_0 &= N(\theta'-\theta)\frac{1}{b} \cdot E\left[\sum_{i=1}^b x_i-\theta-\frac{\theta'-\theta}{2}\right] \\
    &= -N(\theta'-\theta)\left(\theta-0.5+\frac{\theta'-\theta}{2}\right).
\end{align}
Next, consider the minibatch acceptance test $\Lambda^*(\theta,\theta') \not\approx
\psi(u,\theta,\theta')$ required by these methods, where $\not\approx$
informally means ``significantly different from'' under the distribution over
samples of $x_i$. This turns out to be
\begin{align}
\Lambda^*(\theta,\theta') \hspace{0.1in} \not\approx \hspace{0.1in}
\psi(u_0,\theta,\theta') &\iff N(\theta'-\theta) \cdot \frac{1}{b}\sum_{i=1}^b x_i-\theta-\frac{\theta'-\theta}{2} \hspace{0.1in} \not\approx \hspace{0.1in} \log u_0 \\
&\iff \frac{1}{b}\sum_{i=1}^b x_i-\left(\theta+\frac{\theta'-\theta}{2} + \frac{\log u_0}{N(\theta'-\theta)}\right) \hspace{0.1in} \not\approx \hspace{0.1in} 0 \\
&\iff \frac{1}{b}\sum_{i=1}^b x_i-0.5 \hspace{0.1in} \not\approx \hspace{0.1in} 0. \label{eq:accept_test_zero}
\end{align}
Since the $x_i$ have mean 0.5, the resulting test with our chosen $u_0$ will
never correctly succeed and must use all $N$ data points.  Furthermore, if we
sample values of $u$ near enough to $u_0$, the terms in parenthesis will not be
sufficiently different from 0.5 to allow the test to succeed.

The choices above for $\theta$ and $\theta'$ guarantee that
\begin{equation}\label{eq:log_uo_range}
    \log u_0 \in -[0.5,1][0.75,1.5] = [-1.5, -0.375]
\end{equation}
and consider the range of $u$ values
\begin{equation}\label{eq:log_u_range}
    \log u \in \log u_0 + [-0.5,0.375]
\end{equation}
the size of the range in $u$ is at least $\exp([-2,-1.125]) \approx
[0.13534,0.32465]$ and occurs with probability at least $p_2=0.18932$. With $u$
in this range, we rewrite the test as:
\begin{equation}\label{eq:accept_test_rewritten}
    \frac{1}{b}\sum_{i=1}^b x_i-0.5 \hspace{0.1in} \not\approx \hspace{0.1in} \frac{\log u/u_0}{N(\theta'-\theta)}
\end{equation}
so that, as in Equation~\ref{eq:accept_test_zero}, the LHS has expected value
zero.  Given our choice of intervals for the variables, the RHS is in the range
$\frac{1}{\sqrt{N}}[-1,0.75]$. The standard deviation of the LHS given the
interval constraints is at least $0.5/\sqrt{b}$.  And so the gap between the LHS
and RHS in Equation~\ref{eq:accept_test_rewritten} is at most $2\sqrt{b/N}$
standard deviations, limiting the range in which the test will be able to
``succeed'' without requiring more samples.

The samples $\theta$, $\theta'$ and $u$ are drawn independently and so the
probability of the conjunction of these events is $c = p_0 p_1 p_2 = 0.0085$.
